---
title: "Plasmodium falciparum Infection Duration Analysis from ama1 Amplicon Sequencing Data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

Data Preparation
Load and format the raw data which contains the raw read counts per haplotype per timepoint for each individual in the study (data_raw.csv), which corresponds to Supplementary Table 6.

```{r}

library(tidyverse)

# Read the raw data from CSV file
df <- read.csv("data_raw.csv", header = TRUE, sep = ",")

# Perform necessary data formatting and clean-up
df <- df[order(df$Date),]
original_df <- df
df$count <- 1
# Additional data cleaning steps...

```

## Here we will start with the dataset described as supplementary table 6 

```{r}

df <- read.csv("data_raw.csv", header = TRUE, sep = ",") ## This is the same as supplementary table 6. It is important to have the haplotype column (cluster_name), timepoint and ID for each kid in the cohort. 

```

Here make some edits to the table so that it is formatted 

```{r}

df<-df[order(df$Date),] ## order by date
original_df<-df ## to be used after for-loop
df$count<-1 # make column with just 1s to be able to count the number of times a haplotype appears

original_unique<-df ## this is for the following plot of all unique haplotypes across all kids/timepoints

df$Timepoint<-gsub("May 12",-1,df$Timepoint) ## timepoint -1 is May12, we will remove
df$Timepoint<-gsub("May 13",13,df$Timepoint) ## timepoint 13 is May13
df$cluster_name<-gsub("_","",df$cluster_name)


df$Sample<-paste(df$ID,df$cluster_name,sep="_")
df$ID<-NULL # delete kid column
df$cluster_name<-NULL
df$Sample<-gsub("\\.","",df$Sample)
df <- df %>% filter(Timepoint!="MAL") %>%
  distinct(Sample,Timepoint,count) %>%
  spread(Sample,count)
  
df_sorted <- df[order(as.numeric(as.character(df$Timepoint))), ] ## order dates as numeric rather than 1, 10, 11 (this is something R does weirdly)
df_sorted[is.na(df_sorted)] <- 0 ## Table fills missing values with NA but I want to change them to 0

```


For loop to read within the chunks and allow for a certain number of skips in between positive timepoints. Adjust the number of skips by using max_zeroes

```{r}

df_length<-list()


## Make an empty list where I will add the chunks of timepoint when a haplotype is present. Remember to run this empty list before the for loop

#### This is a for loop that finds in the matrix of 0 and 1, chunks of 1s allowing TWO or FOUR SKIPS ####

max_zeroes = 2 ## Here add the number of zeros corresponding to the desired number of skips in-between timepoints

for (name in colnames(df_sorted[,-1])){
  values = df_sorted[[name]]
  # start/end of a chunk
  start<-NA
  end<-NA
  # how many zeroes encountered since
  # the start or a chunk
  num_zeroes<-0
  # container for start/end
  chunks<-list()
  # signpost for last seen 1
  last_one_seen<-NA
  
  for (index in 1:length(values)){
   # print(c(name, index, length(values)))
    if (index!= length(values)){
      
      value=values[index]
      if(value == 0){
        
        # are we inside a chunk?
        if(!is.na(start)){
          
          num_zeroes = num_zeroes + 1
          # have we found a zero already?
          if(num_zeroes > max_zeroes){
            # end of a chunk
            end = index - 1 - max_zeroes
            chunks<-append(chunks,list(c(start, end)))
            
            # reset
            start<-NA
            end<-NA
            num_zeroes<-0
          }
        }
      }
      else {
        # is this the start of a chunk?
        if(is.na(start)){
          start<-index
        }
        # reset the number of zeroes found
        num_zeroes<-0
        last_one_seen<-index
      }
    }
  }
  # handle last data point
  value=values[length(values)]
  # are we inside a chunk?
  if(!is.na(start)){
    # is the last value zero?
    if(value == 0){
      # chunk ends at the last 1 seen
      end = last_one_seen
      chunks<-append(chunks,list(c(start, end)))
    }
    else {
      # chunk ends at last position
      end = length(values)
      chunks<-append(chunks,list(c(start, end)))
      start<-NA
      end<-NA
    }
  }
  else {
    if(value==1){
    #  print(name)
      chunks<-append(chunks,list(c(length(values),length(values))))
    }
  }
  
  x=str_split(name,"_")
  Kid<-x[[1]][1]
  cluster_name<-x[[1]][2]
  for (chunk in chunks){
    start=df_sorted$Timepoint[chunk[1]]
    end=df_sorted$Timepoint[chunk[2]]
    df_length<-append(df_length,list(c(Kid,cluster_name,start,end)))
  }
}


## Now we just make our list containing lists (chunks) into a dataframe with columns kid, haplotype, start and end ##
df_length  <-  as.data.frame(matrix(unlist(df_length), nrow=length(unlist(df_length[1]))))
df_length<-t(df_length)
colnames(df_length)<-c("ID","Haplotype","Start","End")
df_length<-as.data.frame(df_length)

## To be able to map to the original dates, I will first change to the original timepoints, so only changing -1 and 13 for May12 and May13, respectively

df_length$Start<-gsub("13","May13",df_length$Start)
df_length$End<-gsub("-1","May12",df_length$End)
df_length$Start<-gsub("-1","May12",df_length$Start)
df_length$End<-gsub("13","May13",df_length$End)

### Add metadata 

original_unique<-original_df[,c(2,3,6)]
dup<-duplicated(original_unique[,c(1,2,3)])
original_unique<-original_unique[!dup,]
#original_unique$Kid<-gsub("kali","Kali",original_unique$Kid)

### This is more formatting of the data frame and matching to the original dates in the full database form above ###

df_length$kid_start<-paste(df_length$ID,df_length$Start,sep="-")
df_length$kid_end<-paste(df_length$ID,df_length$End,sep="-")

colnames(original_unique)<-c("ID","Timepoint","Date")
original_unique$Sample<-paste(original_unique$ID,original_unique$Timepoint,sep="-")

df_length<-merge(original_unique,df_length,by.x=c("Sample"),by.y=c("kid_start"))
colnames(df_length)<-gsub("Date","start_date",colnames(df_length))
 colnames(df_length)<-gsub("ID","kid_start",colnames(df_length))
df_length<-merge(original_unique,df_length,by.x=c("Sample"),by.y=c("kid_end"))
colnames(df_length)<-gsub("Date","end_date",colnames(df_length))
colnames(df_length)<-gsub("ID","kid_end",colnames(df_length))

df_length$kid_start.x <- NULL
df_length$kid_start.y <- NULL
```

